---
title: "Homework 4"
author: "Nicholas Way"
output:
  html_document:
    df_print: paged
---

##Section 1
a.
Data Tables - tidy data with one or more variables

Arguments - describe the details of what a function is to do

Functions - objects that transform an input to an output; always followed
immediately by an open parenthesis (a few exceptions)

Constants - single values, most commonly a number or character string

Variables - columns in a data table

b. 
== - special type of function used to evaluate equality

%>% - chaining operator

<- - assignment operator

c. 
False - summarise() is a reduction function

True - group_by() should always be followed by another data verb

False - When there is missing data (NA), reduction functions will automatically remove missing data by default and execute the proper calculation with the remaining data

True - ifelse() allows you to translate each value in a variable to one of two values, depending on the result of a comparison

##Section 2
a. The date() function does not take any arguments

b. The date() function results in an object of a character string of the current system date and time

##Section 3

a. Determine the 3rd largest. - Reduction

b. Determine the 3rd and 4th largest values. - Neither

c. Determine the number of cases. - Reduction

d. Determine whether a year is a leap year. - Transformation

e. Determine whether a date is a legal holiday. - Transformation

f. Determine the range of a set, that is, the max minus the min. - Reduction

g. Determine which day of the week (e.g., Sun, Mon, …) a given date is. - Transformation

h. Find the time interval in days spanned by a set of dates. - Transformation

##Section 4
a. The group_by("First") should actually be group_by("name")

b. Replace summarise(Tmp...) with summarise(BabyNames)

c. The code needs the %>%, use n() instead of sum()

##Section 5

```{r}
data("airquality")
head(airquality)
summary(airquality)
```
a. There are 6 different variables within the dataset "airquality". There are 153 cases within the dataset as well. I used two commands. head() to see the first couple of rows of the data, and summary() to just get the summary statistics for each variable within the set.

```{r}
mean(airquality$Temp)
```
b. I used the summarise function of mean(airquality$Temo) to find the mean temperature of 77.88235.

c. 
```{r}
airquality %>%
  group_by(Month) %>%
  summarise(days = n_distinct(Day), temp = median(Temp), minWind = min(Wind), maxWind = max(Wind))
```
















