---
title: "HW 1"
author: "Nicholas Way"
date: "2023-01-24"
output: html_document
---
#Section 1


```{r}
1+2*(3+4)
```

```{r}
(4^3)+(3^2)
```
```{r}
sqrt((4+3)+(2+1))
```
```{r}
(((1+2)/(3+4)))^2
```
#Section 2

```{r}
rep("a", 4)
```
```{r}
 seq(from=1, to=100, by=2)
```
```{r}
rep(2:4, c(3,3,3))
```
```{r}
rep(2:4, 2)
```
```{r}
seq(from=0, to=1000, by=25 )
```

#Section 3

```{r}
x1 <- c(1.7, "a", FALSE)
x2 <- c(TRUE, 2)
```
```{r}
class(x1)
```
```{r}
class(x2)
```


#Section 4

1. "essay14" is a legitimate object name because it does not break any of the rules when it comes to naming objects.

2. "first-essay" is not a legitimate object name because it breaks a rule by using the "-" symbol.

3. ""MyData"" is not a legitimate object name because it contains the symbols "" which are not allowed.

4. "small sample is not a legitimate object name because it contains a space in between the two words.

5. ".myData" is a legitimate object name because it doesn't break any of the naming objects rules.

6. "sqrt()" is not a legitimate object name because it contains the symbols "()".

#Section 5

```{r}
bill <- c(46, 33, 39, 37, 46, 30, 48, 32, 49, 35, 30, 48)
```

```{r}
bill[2:12]
```























