---
title: "HW1"
author: "Nicholas Way"
date: "2023-08-24"
output: html_document
---

## Front Matter
```{r}
data(iris)
library(tidyverse)
```

## Problem 1
```{r}
ggplot(data = iris, mapping = aes(x = Sepal.Length,))+
  geom_histogram(binwidth = .25, color = "black", fill = "red")+
  labs(x = "Sepal Length (in cm)",
       y = "Number of Flowers",)
```

- The shape of the distribution of the number of flowers with varying sepal lengths is fairly symmetrical while being slightly skewed right.

## Problem 2
```{r}
ggplot(data = iris, mapping = aes(x = Species)) +
  geom_bar(fill = "white", color = "black") +
  labs(x = "Species",
       y = "Number of Flowers",)
```

- The distribution of the number of flowers for each species is completely identical. There is exactly 50 flowers within the data set for each of the three species.

## Problem 3
```{r}
ggplot(data = iris, mapping = aes(x = Sepal.Length, y = Sepal.Width))+
  geom_point(shape = 6)+
  geom_smooth(method = lm, se = FALSE) +
  labs(x = "Sepal Length (in cm)",
       y = "Sepal Width (in cm)",)
```

- After analyzing the scatterplot with regression line, I have learned that there appears to be a very weak negative correlation between sepal length and sepal width, meaning that as the sepal length of a flower increases, it is expected that the sepal width decreases. This is not what I expected, I expected the variables to move in the same direction, meaning that as length increases, width increases.

## Problem 4
```{r}
ggplot(data = iris, mapping = aes(x = Sepal.Length, y = Sepal.Width, color = as.factor(Species)))+
  geom_point()+
  geom_smooth(method = lm, se = FALSE) +
  labs(x = "Sepal Length (in cm)",
       y = "Sepal Width (in cm)",
       color = "Species")
```

- After categorizing the relationship between length and width by species type, I have learned that the species versicolor and virginica both have a slightly positive correlation between their sepal length and width, while the species setosa has a much larger positively correlated relationship between their sepal length and width. It appears that no matter the species, as sepal length increases, so does sepal width. This is what I expected

## Problem 5
```{r}
sp <- ggplot(data = iris, mapping = aes(x = Sepal.Length, y = Sepal.Width, color = as.factor(Species)))+
  geom_point()+
  geom_smooth(method = lm, se = FALSE) +
  labs(x = "Sepal Length (in cm)",
       y = "Sepal Width (in cm)",
       color = "Species")

sp2 <- sp + scale_color_manual(values=c("#E69F00", "#39FF14", "#00008B"))
sp2
```

- Google Search : how to change the color scheme from the default colors to orange, green, and navy in r ggplot

- Google Search : color code in r for dark blue

- Link : http://www.sthda.com/english/wiki/ggplot2-colors-how-to-change-colors-automatically-and-manually

## Problem 6
```{r}
sp2 + theme(legend.position = "bottom")
```

- Google Search : how to change the location of the legend in r ggplot

- Link : https://r-graph-gallery.com/239-custom-layout-legend-ggplot2.html#:~:text=You%20can%20place%20the%20legend,the%20x%20and%20y%20coordinates.

## Problem 7
```{r}
ggplot(data = iris, mapping = aes(x = Sepal.Length, y = Sepal.Width,))+
  geom_point()+
  geom_smooth(method = lm,) +
  theme_bw() +
  facet_grid(. ~ Species) +
  labs(x = "Sepal Length (in cm)",
       y = "Sepal Width (in cm)",)
```

- Google Search : how to remove gray background in r ggplot

- Link : https://felixfan.github.io/ggplot2-remove-grid-background-margin/

## Problem 8 
```{r}
ggplot(data = iris, mapping = aes(x = Species, y = Sepal.Width)) +
  geom_boxplot() +
  labs(x = "Species", 
       y = "Sepal Width (in cm)") +
  theme_bw()
```















